---
title: "Analyzing ancient DNA data with tfa" 
author: "Olivier François - Flora Jay" 
date: "`r Sys.Date()`" 
output: 
  rmarkdown::html_vignette
  #pdf_document:
  #  latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**Summary:** The R package `tfa` performs factor analyses of temporal or ancient DNA samples, correcting individual scores for the effect of allele frequency drift through time. The package implements a fast algorithm that computes $K$ corrected factors, and provides estimates of ancestry proportions for a target individual or population given a  set of source populations.


```{r setup}
library(tfa)
```


#### Main function 

```{r}
library(tfa)

# Ancient DNA from Bronze Age Great Britain samples
# including Yamnaya, early farmers (Anatolia) and hunter-gatherers (Serbia)
data(england_ba)

attach(England_BA)
coverage <- meta$Coverage
geno <- coverage_adjust(genotype, coverage, K = 3, log = TRUE)

mod  <- tfa(age,
            geno,
            k = 2,
            lambda = 5e-1,
            center = TRUE,
            coverage = coverage,
            log = TRUE)

plot(mod$u, pch = 19, cex = 2, col = "grey90",
     xlab = "Factor 1", ylab = "Factor 2",
     main = "FA")

m_yamnaya <- apply(mod$u[meta$Group.ID == "Russia_Yamnaya",],
                   2, mean)
m_anatolia <- apply(mod$u[meta$Group.ID == "Anatolia_N",],
                    2, mean)
m_hg <- apply(mod$u[meta$Group.ID == "Serbia_HG",],
              2, mean)

points(rbind(m_yamnaya, m_anatolia, m_hg), lwd = 2)

lines(rbind(m_yamnaya, m_anatolia, m_hg, m_yamnaya))

points(mod$u[meta$Group.ID == "Russia_Yamnaya",],
       pch = 8, cex = .6, col = "darkblue")

points(mod$u[meta$Group.ID == "Anatolia_N",],
       pch = 8, cex = .6, col = "salmon3")

points(mod$u[meta$Group.ID ==  "Serbia_HG",],
       pch = 8, cex = .6, col = "olivedrab")

points(mod$u[meta$Group.ID == "England_Bell_Beaker",],
       pch = 19, cex = .6, col = "yellow4")

points(mod$u[meta$Group.ID == "England_BA",],
       pch = 19, cex = .6, col = "yellow3")

points(mod$u[meta$Group.ID %in% c("England_N", "Scotland_N"),],
       pch = 19, cex = .6, col = "salmon1")

legend(x = 10, y = -15, cex = .6,
       legend = c("Early Farmers", "Hunter Gatherers", "Steppe"),
       col = c("salmon3", "olivedrab", "darkblue"), pch = 8)

legend(x = 10, y = 20.5, cex = .6,
       legend = c("Neolithic GB", "Bronze Age GB", "Bell Beaker"),
       col = c("salmon1", "yellow3", "yellow4"), pch = 19)
detach(England_BA)
# rm(list = ls())
```


#### Ancestry coefficients

```{r}
# Ancient DNA from Bronze Age Great Britain samples
# including Steppe (Yamnaya), hunter gatherers, and early farmers from Anatolia
data(england_ba)
attach(England_BA)
metadata <- as.tfa_data(meta)
coverage <- meta$Coverage
geno <- coverage_adjust(genotype, coverage, K = 3, log = TRUE)

mod  <- tfa(age,
            geno,
            k = 3,
            lambda = 5e-1,
            center = TRUE,
            coverage = coverage,
            log = TRUE)

target = c("England_Bell_Beaker", "England_BA")
source = c("Anatolia_N", "Russia_Yamnaya", "Serbia_HG")
ancestry_coefficients(model = mod,
                      metadata = metadata,
                      source = source,
                      target = target)

source = c("Scotland_N", "Russia_Yamnaya", "Serbia_HG")
ancestry_coefficients(model = mod,
                      metadata = metadata,
                      source = source,
                      target = target)
detach(England_BA)
# rm(list = ls())
```



